\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-02-01 docs/papers/sto/so.tex
FIXUP draft updates

Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD), neural
network (NN), particle-mesh (PM), fast-Fourier transform (FFT)

% AI for science and cosmology

@@

% introduce PM and its limitation on small scales
The PM method solves the Poisson equation by scattering the particles onto a
mesh, on which FFT is performed and the Laplacian operator reduces to simple
multiplication of wavenumbers in Fourier space.
Although being computationally efficient, the PM method is mainly limited by its
resolution on small scales, i.e. scales around the mesh cell size.
This is due to the loss of information during the scattering process of
continuously distributed particles onto the discrete mesh.
We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to compute the
mass fractions of a particle scattered to nearby grid points.
Other schemes include the nearest grid point (NGP) and triangle shaped cloud
(TSC) etc.

Using a hybrid embedding of NNs in differentiable physical simulations, we
optimize the performance of PM method on small scales in cosmological $N$-body
simulations by sharpening the gravitational force.

@@

In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
where we introduce force sharpening for spatial optimization through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ summarized
in \autoref{tab:feat} with more details discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

@@

\subsection{Trained Neural Networks}

% visualization & comparison of neural nets & symbolic functions
For the $f(k_i; \vvartheta)$ NN, we show it as a function of $k_i$ at multiple
different $\vvartheta$ values.

For the $g(k_1,k_2,k_3;\vvartheta)$ NN, which is permutationally symmetric about
$(k_1, k_2, k_3)$, we show it as a function of $k_i$ along axis and face
diagonal directions.

\subsection{Feature Importance}

\subsection{Symbolic Expressions}

@@

\begin{figure*}
  \centering
  \includegraphics[width=.8\columnwidth]{sobol0snap120epoch300.pdf}
  \caption{The transfer function $T$ and correlation coefficient $r$ of the
  \GADGET\ and \pmwd\ simulated density fields, with and without force
  sharpening.
  Sobol 0 snap 120}
  \label{fig:tfcc}
\end{figure*}

% transfer function & correlation coefficient
To compare two fields $\hat f$ and $f$, we introduce for each mode the square
root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
which is also referred to as the transfer function, and the cross correlation
coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
\label{r}
\end{equation}
%
The two fields are identical when both $T$ and $r$ are unity.
The PM method is accurate on scales much larger than the cell size, therefore we
expect $T$ and $r$ to be very close to being unity both before and after force
sharpening optimization.

@@

\subsection{Dark Matter Halo Catalogs}

@@

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the gradient-based optimization discussed in this work.
Within this fully differentiable framework, we are able to embed neural networks
in physical processes and train them for many purposes, including the
optimization of spatial resolution by sharpening the gravitational force in this
work.

Here we briefly recap the equations that are closely related to our
extensions to the original framework, and setup the notations of the variables.
Following \cite{Li2022a}, we denote the particles' state with $\vz = (\vx,
\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
We refer to the quantities that are directly related to the final objective
function $\cJ$ as the observable $\vo$.
In this work, our observable includes multiple snapshots at given times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations (check
Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to $a_{i-1}$, the adjoint
variables are updated through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero only if they were used in the generation of any snapshots in $\vo$.
The initial conditions of these adjoint variables are given by
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
The derivative of the objective function with respect to the observable
($\p\cJ/\p\vo$) and the VJPs of the observation operation ($\p\vo/\p\vx_n$ and
$\p\vo/\p\vp_n$, mainly the interpolation function discussed in
\autoref{sec:snapobs} below) can be conveniently computed by AD.

@@

The N-body time integration is carried out following discrete time steps. Given
the particle displacements $\vx$ and velocities $\vv = \vp/(a^3H)$ at two
consecutive time steps $a_i$ and $a_{i+1}$, we get the snapshot of particles in
between at $a$ using
the cubic Hermite interpolation.
The interpolated particle displacements are given by
\begin{align}
  \vx(a) =\ &h_{00}(\alpha)\vx_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
\end{align}
where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis functions read
\begin{align}
  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
\end{align}
The particle velocities are then given by the derivative $\d\vx / \d a$.
Note that the interpolation is a linear combination of the particle snapshots at
two consecutive time steps, which therefore can be done in either two steps or
just one step by carrying the particles from the previous time step.
We find the two-steps scheme more convenient for the backward evolution of the
adjoint variables.
The interpolated snapshots at different times make the observable in this work,
and the VJPs of the interpolation operation are computed by AD and put into the
adjoint equations as mentioned in \autoref{sec:pmwd}.

@@

\vspace{1em}
\subsection{Training of SO Neural Networks}

@@

@@

\vspace{1em}
\subsection{Feature Importance Analysis}

@@

\vspace{1em}
\textit{\large Acknowledgements:}
We thank Qiufan Lin, Paulo Montero-Camacho, and Yang Wang for helpful
discussions.
YZ and YL were supported by The Major Key Project of PCL.
The Flatiron Institute is supported by the Simons Foundation.

@@

\vspace{1em}
\textit{\large Code Availability:}
\pmwd\ \citep{Li2022b} is open-source on GitHub
\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source files
and scripts of this paper
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
\GADGET\ \citep{GADGET-4}, which can be obtained through a public git-repository
\href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab}, hosted by a gitlab
server at the Max-Planck Computing and Data Facility (MPCDF).

>>>>

% AI for science and cosmology
Rapid advances in artificial intelligence (AI) and more specifically
deep learning (DL) have provided powerful tools that are revolutionizing
scientific research across disciplines.
AI for science consists of not only the state-of-art deep neural
networks, but also more fundamental computational tools and hardwares
like automatic differentiation (AD) and GPUs.
One particular application is the numerical simulation of the Universe.

@@

% introduce PM and its limitation on small scales
The particle-mesh (PM) method solves the Poisson equation by scattering
the particles onto a mesh, on which FFT is performed and the Laplacian
operator reduces to simple multiplication of wavenumbers in Fourier
space.
Although being computationally efficient, the PM method is mainly
limited by its resolution on small scales, i.e. scales around the mesh
cell size.
This is due to the loss of information during the scattering process of
continuously distributed particles onto the discrete mesh.
We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to
compute the mass fractions of a particle scattered to nearby grid
points.
Other schemes include the nearest grid point (NGP) and triangle shaped
cloud (TSC) etc.

% motivation of this work
Optimizing the accuracy of fast and differentiable simulations is
crucial for their applications in, e.g. generation of mocks for
cosmological surveys, simulation-based inference or field-level
inference.

% introduction of this work
The differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b} lay the foundation of the gradient-based optimization
discussed in this work.
Within this fully differentiable framework, we sharpen the gravitational
force by embedding neural networks (NNs) in the Poisson equation and
train them for the spatial optimization (SO) of the simulation
resolution on small scales.
For better interpretability, we further fit the trained NNs using
symbolic regression (SR).

@@

In cosmological $N$-body simulations, the dynamics of the particles is
governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\delta(\vx)$ and $\varphi(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
where for each dimension $i\in\{1,2,3\}$ of the 3D field, we then
introduce SO force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$ summarized in \autoref{tab:feat} with more details
discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.

Being embedded in the differentiable simulation, the NNs can be trained
using gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation
in every time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron
(MLP, i.e. fully connected) structures.

@@

\subsection{Trained Neural Networks \& Symbolic Expressions}

\begin{figure*}
  \centering
  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
  functions }
  \label{fig:func}
\end{figure*}

With discrete Fourier transform (DFT) on a regular mesh of length $N_c$
with cell size $l_c$, the range of sampled 1D discrete wavenumbers is
given by the fundamental $k_{\rm fund} = 2\pi/(l_c N_c)$ and the
maximum $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is simply the box
size.
To visualize the functions for different box sizes, we normzalize $k$ by
$k_{\rm Nyquist}$, which gives the same $[2/N_c, 1]$ regardless of the
box size.
For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum value
is then $\sqrt{3}$.

For the $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally
symmetric about $(k_1, k_2, k_3)$, we show it as a function of $k_i$
along axis, face and body diagonal directions.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.48\columnwidth]{tf_snap120_3091776_e3000.pdf}
  \includegraphics[width=.48\columnwidth]{cc_snap120_3091776_e3000.pdf}
  \caption{The TF and CC of PM and PM-SONN simulated density fields with
  respect to that of the \GADGET. The solid line and shaded region
  indicate the median and the [16, 84] percentiles of the 64 Sobol
  simulation setups. last snapshot $a\simeq 1$, mesh shape=1}
  \label{fig:tfcc}
\end{figure*}

To compare two fields $\hat f$ and $f$, we introduce for each mode the
square root ratio of the auto power spectrum
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
which is referred to as the transfer function (TF), and the correlation
coefficient (CC) defined with the cross power spectrum
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
\label{r}
\end{equation}
%
The two fields are identical when both $T$ and $r$ are unity.
The PM method is accurate on scales much larger than the mesh cell size,
therefore we expect $T$ and $r$ to be very close to being unity both
before and after the force sharpening optimization.
To show the $T$ and $r$ for different box sizes, we normalize the
wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
the particle grid, where $l_p$ is the particle spacing.
\YZ{We show T and r separately for different mesh shapes?}

comparison of PM, PM with SO NN (PM-SONN), PM with SO SR (PM-SOSR)

@@

\subsection{Dark Matter Halos}

\subsection{Benchmarks}
We compare the computational cost of \GADGET, \pmwd\ with NNs, and
\pmwd\ with SR.

@@

The spatial optimization using gradient descent discussed in this work
is built in the framework of the PM $N$-body library \pmwd\
\citep{Li2022b}, which is an implementation of the differentiable
cosmological simulation with adjoint method \citep{Li2022a}.

Here we briefly recap the equations that are closely related to our
extensions of the original framework, and introduce the notations of the
variables.
Following \cite{Li2022a}, we denote the particles' state with $\vz =
(\vx, \vp)^\intercal$, where $\vx$ and $\vp$ are the positions and
momenta of the particles, and their adjoint variables with $\vlambda =
(\vxi, \vpi)$.
We refer to the quantities that are directly related to the final
objective function $\cJ$ as the observable $\vo$.
In this work, our observables include multiple snapshots at given times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations
(check Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to
$a_{i-1}$, the adjoint variables are updated through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero only if they were used in the generation of any snapshots in $\vo$.
The initial conditions of these adjoint variables are given by
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$, the
state of particles at the final time step.
The derivative of the objective function with respect to the observable
($\p\cJ/\p\vo$) and the VJPs of the observation operation
($\p\vo/\p\vx_n$ and $\p\vo/\p\vp_n$, mainly the interpolation function
discussed in \autoref{sec:snapobs} below) can be conveniently computed
by AD.

@@

The N-body time integration is carried out along discrete time steps in
terms of the scale factor $a$, and the state of particles at any given
time can be achieved using interpolation.
Given the particle displacements $\vs$ and velocities\footnote{This
velocity is related to the canonical momentum $\vp$ through $\vv =
\vp/(a^3H)$.} $\vv := \d\vs/\d a$ at two consecutive time steps $a_i$ and
$a_{i+1}$, we get the snapshot of particles in between at $a$ using the
cubic Hermite interpolation.
The interpolated particle displacements are given by
\begin{align}
  \vs(a) =\ &h_{00}(\alpha)\vs_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
           &h_{01}(\alpha)\vs_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
\end{align}
where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis
functions read
\begin{align}
  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
\end{align}
The particle velocities are then given by the derivative $\vv(a) =
\d\vs(a) / \d a$.
The interpolation is a linear combination of the particle snapshots at
two consecutive time steps, which therefore can be done in either two
steps or just one step by carrying the particles from the previous step.
We find the two-steps scheme more convenient for the backward evolution
of the adjoint variables.
The interpolated snapshots at different times make the observable in
this work, and the VJPs of the interpolation functionx are computed by
AD and put into the adjoint equations as mentioned in
\autoref{sec:pmwd}.

@@

\vspace{1em}
\subsection{Training of Neural Networks}

@@

Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
A rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply an exponential function to the output as a regulator
for positive output.

@@

@@

\vspace{1em}
\textit{\large Acknowledgements:}
We thank Qiufan Lin, Paulo Montero-Camacho, and Yang Wang for helpful
discussions.
YZ and YL were supported by The Major Key Project of PCL.
YZ was also supported by the China Postdoctoral Science Foundation under award
number 2023M731831.
The Flatiron Institute is supported by the Simons Foundation.

@@

\vspace{1em}
\textit{\large Code Availability:}
\pmwd\ \citep{Li2022b} is open-source on GitHub
\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
files and scripts of this paper
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
\GADGET\ \citep{GADGET-4} can be obtained through a public
git-repository \href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab},
hosted by a gitlab server at the Max-Planck Computing and Data Facility
(MPCDF).
